---
title: "French MTPL GLM Walkthrough"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

source("mtpl_glm_analysis.R")
results <- run_mtpl_analysis()
checks <- results$checks
flags <- results$flags
holdout_summary <- results$holdout_summary
example_policies <- results$example_policies
gas_summary <- results$gas_summary
age_summary <- results$age_summary

get_check <- function(name) {
  unname(checks$value[checks$metric == name])
}

format_eur <- function(x) {
  paste0("EUR ", format(round(x, 0), big.mark = ",", scientific = FALSE))
}

low_freq <- example_policies["Lower-risk example", "pred_frequency_annual"]
low_sev <- example_policies["Lower-risk example", "pred_severity"]
low_premium <- example_policies["Lower-risk example", "pred_pure_premium"]

typical_freq <- example_policies["Typical example", "pred_frequency_annual"]
typical_sev <- example_policies["Typical example", "pred_severity"]
typical_premium <- example_policies["Typical example", "pred_pure_premium"]

high_freq <- example_policies["Higher-risk example", "pred_frequency_annual"]
high_sev <- example_policies["Higher-risk example", "pred_severity"]
high_premium <- example_policies["Higher-risk example", "pred_pure_premium"]

diesel_premium <- gas_summary$pred_pure_premium[gas_summary$Gas == "Diesel"]
regular_premium <- gas_summary$pred_pure_premium[gas_summary$Gas == "Regular"]

young_premium <- age_summary$pred_pure_premium[age_summary$DriverAgeBand == "[17,25]"]
old_premium <- age_summary$pred_pure_premium[age_summary$DriverAgeBand == "(70,Inf]"]

practical_comparison <- data.frame(
  comparison = c(
    "Higher-risk vs lower-risk example",
    "Age 17-25 vs age 70+",
    "Diesel vs Regular"
  ),
  ratio = c(
    high_premium / low_premium,
    young_premium / old_premium,
    diesel_premium / regular_premium
  ),
  row.names = NULL
)
```

# Executive Results

I wrote this notebook in the same order I actually worked through the project. I started by loading the data, then I checked whether it looked sane, then I fit a simple baseline, then I dealt with the parts that looked off, and finally I translated the results into a pricing story that feels intuitive.

The headline results are:

- average predicted annual claim frequency is `r sprintf("%.2f%%", 100 * mean(results$pure_premium$pred_frequency_annual))`
- average predicted claim severity is `r format_eur(mean(results$pure_premium$pred_severity))`
- average predicted pure premium is `r format_eur(mean(results$pure_premium$pred_pure_premium))` per policy-year
- holdout aggregate frequency is very close: `r round(get_check("holdout_observed_claims"), 0)` observed claims versus `r round(get_check("holdout_predicted_claims"), 1)` predicted
- raw severity underpredicts holdout mean severity by about `r sprintf("%.1f%%", 100 * (get_check("holdout_predicted_mean_severity") / get_check("holdout_observed_mean_severity") - 1))`
- capped severity sensitivity is much tighter: `r format_eur(get_check("holdout_observed_capped_mean_severity"))` observed versus `r format_eur(get_check("holdout_predicted_capped_mean_severity"))` predicted

At a high level, what I found feels pretty intuitive. Younger drivers and some higher-frequency segments come out as more expensive. The severity side is a little messier because the claim amounts have a heavy tail, so I tried to be honest about that instead of pretending the model was cleaner than it really is.

# Step 1: Load The Data And Confirm The Shape

I started with the simplest possible base-`R` import.

```{r step1-code, eval=FALSE}
freq <- read.csv("freMTPLfreq.csv", stringsAsFactors = FALSE)
sev <- read.csv("freMTPLsev.csv", stringsAsFactors = FALSE)

nrow(freq)
nrow(sev)
```

Then I checked that the dataset size matched what I expected after loading.

```{r step1-results}
checks[checks$metric %in% c("policy_rows_raw", "claim_rows_raw"), ]
```

What I took from this:

- This looked right immediately.
- I have `r format(get_check("policy_rows_raw"), big.mark = ",", scientific = FALSE)` policy rows and `r format(get_check("claim_rows_raw"), big.mark = ",", scientific = FALSE)` claim rows.
- That lines up with the dataset description, so I did not see any sign of a bad import, broken delimiter, or accidental filtering.

# Step 2: Clean Only What We Need

I wanted this to feel like a believable actuarial baseline, not a giant preprocessing project, so I kept the cleaning deliberately light.

```{r step2-code, eval=FALSE}
freq <- freq[complete.cases(freq[, c(
  "PolicyID", "ClaimNb", "Exposure", "Power", "CarAge",
  "DriverAge", "Brand", "Gas", "Region", "Density"
)]), ]
freq <- freq[freq$Exposure > 0, ]
freq$logDensity <- log1p(freq$Density)

sev <- sev[complete.cases(sev[, c("PolicyID", "ClaimAmount")]), ]
sev <- sev[sev$ClaimAmount > 0, ]
```

```{r step2-results}
checks[checks$metric %in% c(
  "policy_rows_raw",
  "policy_rows_after_cleaning",
  "claim_rows_raw",
  "claim_rows_after_cleaning"
), ]
```

What I took from this:

- Nothing alarming happened here.
- I did not lose any rows in the light cleaning pass, which is a good sign for a clean benchmark dataset.
- I log-transformed `Density` because raw population density is highly skewed and a plain linear effect would be hard to defend.

# Step 3: Build A Train/Holdout Split

Before fitting anything, I created one fixed train/holdout split at the policy level, stratified on whether the policy had any claims.

```{r step3-code, eval=FALSE}
train_flag <- stratified_split(as.integer(freq$ClaimNb > 0), seed = 20260423)
freq$split <- ifelse(train_flag, "train", "holdout")

sev_model_data <- merge(
  sev,
  freq[, c("PolicyID", "split", "Power", "CarAge", "DriverAge",
           "Brand", "Gas", "Region", "Density", "logDensity")],
  by = "PolicyID",
  sort = FALSE
)
```

```{r step3-results}
checks[checks$metric %in% c(
  "train_policies",
  "holdout_policies",
  "train_claim_rows",
  "holdout_claim_rows"
), ]
```

What I took from this:

- This also looked fine.
- The split keeps a large training sample and a meaningful holdout sample.
- Most importantly, claims stay attached to the same policy split, so there is no leakage from a policy's claims into both training and validation.

# Step 4: Fit The First Frequency Model

For the first frequency model, I went with the classical starting point: a Poisson GLM with a log link and an exposure offset.

```{r step4-code, eval=FALSE}
freq_fit <- glm(
  ClaimNb ~ Power + CarAge + DriverAge + Brand + Gas + Region + logDensity,
  family = poisson(link = "log"),
  data = freq_train,
  offset = log(Exposure)
)
```

Then I checked whether it was at least directionally right on unseen data.

```{r step4-results}
holdout_summary[holdout_summary$metric %in% c(
  "Observed claims",
  "Predicted claims",
  "Observed annual frequency",
  "Predicted annual frequency",
  "Frequency dispersion"
), ]
```

```{r step4-frequency-plot}
freq_deciles <- results$frequency_deciles
plot(
  x = seq_len(nrow(freq_deciles)),
  y = freq_deciles$observed_frequency,
  type = "b",
  pch = 19,
  xaxt = "n",
  xlab = "Holdout annual predicted-frequency decile",
  ylab = "Annual frequency",
  main = "Frequency calibration on holdout"
)
axis(1, at = seq_len(nrow(freq_deciles)), labels = freq_deciles$decile, las = 2, cex.axis = 0.7)
lines(seq_len(nrow(freq_deciles)), freq_deciles$predicted_frequency, type = "b", pch = 17, col = "firebrick")
legend("topleft", legend = c("Observed", "Predicted"), col = c("black", "firebrick"), lty = 1, pch = c(19, 17), bty = "n")
```

What I took from this:

- The good news is that aggregate holdout frequency is excellent: `r round(get_check("holdout_observed_claims"), 0)` observed versus `r round(get_check("holdout_predicted_claims"), 1)` predicted.
- The less good news is that dispersion is `r round(get_check("frequency_dispersion"), 2)`, which is too high for a clean Poisson inference story.
- So my basic frequency signal looked useful, but the plain Poisson standard errors were too optimistic.
- To me, that felt like a normal insurance-modeling issue, not a sign that the whole model was failing.

# Step 5: Fix Frequency Inference With Quasi-Poisson

I did not want to overreact and blow up the scope. The narrow fix was to keep the Poisson mean structure for prediction and use quasi-Poisson as the more honest inference check.

```{r step5-code, eval=FALSE}
freq_quasi_fit <- glm(
  ClaimNb ~ Power + CarAge + DriverAge + Brand + Gas + Region + logDensity,
  family = quasipoisson(link = "log"),
  data = freq_train,
  offset = log(Exposure)
)
```

```{r step5-results}
round(results$top_frequency_effects_quasi, 4)
```

What I took from this:

- This is where the frequency story started to feel more credible.
- The main directional findings survived the quasi-Poisson adjustment, but some marginal effects became less certain once I acknowledged the extra variance.
- The most stable takeaway still feels practical rather than theoretical:
  - Regular fuel has lower predicted claim frequency than Diesel in this dataset.
  - Some Japanese/Korean brand groups show lower claim frequency.
  - Several higher power levels carry higher expected frequency.
- I do not read these as causal claims. I read them as pricing signals.

# Step 6: Fit The First Severity Model

For severity, I stuck with the classical simple choice: a Gamma GLM with a log link on positive claim amounts.

```{r step6-code, eval=FALSE}
sev_fit <- glm(
  ClaimAmount ~ Power + CarAge + DriverAge + Brand + Gas + Region + logDensity,
  family = Gamma(link = "log"),
  data = sev_train
)
```

```{r step6-results}
holdout_summary[holdout_summary$metric %in% c(
  "Observed mean severity",
  "Predicted mean severity"
), ]
```

```{r step6-severity-effects}
round(results$top_severity_effects, 4)
```

What I took from this:

- This is where the first baseline started to feel less comfortable.
- Holdout mean severity is `r format_eur(get_check("holdout_observed_mean_severity"))`, while the raw Gamma model predicts `r format_eur(get_check("holdout_predicted_mean_severity"))`.
- That is not catastrophic, but it is weak enough that I do not want to pretend the raw severity model is perfectly calibrated.
- The coefficient table also looks noisy, which is exactly what I would expect when a heavy-tailed severity distribution is being fit with a simple GLM.

# Step 7: Check The Tail And Add A Capped-Loss Sensitivity

Instead of throwing the model away, I tried to ask a better question: is the model bad everywhere, or is it mostly getting pulled around by a handful of very large losses?

The simplest stock-`R` sensitivity was to cap training severities at the 99.5th percentile and refit the Gamma model.

```{r step7-code, eval=FALSE}
severity_cap <- as.numeric(quantile(sev_train$ClaimAmount, probs = 0.995))
sev_train$ClaimAmountCapped <- pmin(sev_train$ClaimAmount, severity_cap)
sev_holdout$ClaimAmountCapped <- pmin(sev_holdout$ClaimAmount, severity_cap)

sev_capped_fit <- glm(
  ClaimAmountCapped ~ Power + CarAge + DriverAge + Brand + Gas + Region + logDensity,
  family = Gamma(link = "log"),
  data = sev_train
)
```

```{r step7-results}
holdout_summary[holdout_summary$metric %in% c(
  "Observed mean severity",
  "Predicted mean severity",
  "Observed capped mean severity",
  "Predicted capped mean severity"
), ]
```

```{r step7-cap}
results$severity_cap
```

```{r step7-influence}
head(round(results$severity_cooks_top, 4), 10)
```

```{r step7-capped-effects}
round(results$top_severity_effects_capped, 4)
```

What I took from this:

- This step told me the real story.
- The cap is `r format_eur(results$severity_cap)`.
- Raw holdout severity is off, but capped holdout severity is much tighter: `r format_eur(get_check("holdout_observed_capped_mean_severity"))` observed versus `r format_eur(get_check("holdout_predicted_capped_mean_severity"))` predicted.
- That tells me the model is doing a more reasonable job on the body of the loss distribution than the raw mean alone suggests.
- In practice, I would describe the severity issue as tail sensitivity, not as a totally useless model.

# Step 8: Combine Frequency And Severity Into Pure Premium

Once I had the two pieces fit, I annualized expected frequency and multiplied by expected severity.

```{r step8-code, eval=FALSE}
annualized_policies$Exposure <- 1
annualized_policies$pred_frequency_annual <- predict(freq_fit, newdata = annualized_policies, type = "response")
annualized_policies$pred_severity <- predict(sev_fit, newdata = annualized_policies, type = "response")
annualized_policies$pred_pure_premium <- annualized_policies$pred_frequency_annual * annualized_policies$pred_severity
```

```{r step8-quantiles}
round(quantile(results$pure_premium$pred_pure_premium, probs = c(0.1, 0.25, 0.5, 0.75, 0.9, 0.95)), 2)
```

```{r step8-hist}
hist(
  results$pure_premium$pred_pure_premium,
  breaks = 40,
  main = "Predicted annual pure premium",
  xlab = "Expected claim cost per policy-year",
  col = "gray80",
  border = "white"
)
```

What I took from this:

- The average pure premium is `r format_eur(mean(results$pure_premium$pred_pure_premium))` per policy-year.
- The 10th, 50th, and 90th percentiles are roughly `r format_eur(quantile(results$pure_premium$pred_pure_premium, 0.10))`, `r format_eur(quantile(results$pure_premium$pred_pure_premium, 0.50))`, and `r format_eur(quantile(results$pure_premium$pred_pure_premium, 0.90))`.
- That spread feels meaningful, but not so extreme that the model looks unhinged.

# Step 9: Check Whether The Segmentation Makes Practical Sense

This is the part I care about most. A model can fit okay statistically and still tell a pricing story that feels silly. So I looked at a few segment summaries and example policyholders to see whether the outputs made intuitive sense.

## Age And Fuel Type Summaries

```{r step9-age}
age_summary
```

```{r step9-gas}
gas_summary
```

What I took from this:

- Driver age behaves the way I would expect.
- The model predicts pure premium of `r format_eur(young_premium)` for ages 17-25 and `r format_eur(old_premium)` for ages 70+, which is a ratio of `r round(young_premium / old_premium, 2)` to 1.
- That feels intuitive to me: younger drivers tend to have higher claim frequency, and in this model they also carry somewhat higher severity.
- Fuel type shows a smaller but still meaningful difference: Diesel is about `r round(diesel_premium / regular_premium, 2)` times the pure premium of Regular.
- I would not tell a causal story there. My guess is that fuel type is acting as a proxy for vehicle usage or mix.

## Example Policyholders

```{r step9-examples}
example_display <- example_policies
numeric_cols <- vapply(example_display, is.numeric, logical(1))
example_display[numeric_cols] <- round(example_display[numeric_cols], 3)
example_display
```

```{r step9-ratios}
practical_comparison
```

What I took from this:

- The lower-risk example has predicted pure premium `r format_eur(low_premium)`.
- The typical example has predicted pure premium `r format_eur(typical_premium)`.
- The higher-risk example has predicted pure premium `r format_eur(high_premium)`.
- The higher-risk example is about `r round(high_premium / low_premium, 2)` times the lower-risk example.

Why does that happen in practice?

- The higher-risk example is worse on frequency and severity at the same time.
- Its annual predicted frequency is `r sprintf("%.2f%%", 100 * high_freq)` versus `r sprintf("%.2f%%", 100 * low_freq)` for the lower-risk example.
- Its predicted severity is `r format_eur(high_sev)` versus `r format_eur(low_sev)`.
- So it is not "paying more" because of one weird coefficient. It is paying more because the model expects both more claims and more expensive claims.

Why does that broadly make sense?

- The higher-risk example is younger.
- It sits in a region with a higher modeled frequency profile.
- Its brand group and overall feature mix push severity upward as well.
- The lower-risk example, by contrast, has an older driver, lower expected frequency, and lower expected severity.

That is the way I would explain this to a non-technical person: not by reciting coefficient tables, but by showing how predicted claim count and predicted claim size combine into a higher expected annual cost.

# What Checked Out, What Did Not, And What I Changed

What checked out for me:

- dataset dimensions matched expectations
- light cleaning did not expose data-quality problems
- holdout aggregate frequency was very strong
- age segmentation makes intuitive sense
- pure premium spread across policies looks believable

What did not check out cleanly:

- Poisson dispersion was too high for a naive inference story
- raw Gamma severity underpredicted holdout mean severity
- raw severity coefficients looked noisy because of the heavy tail

What I changed:

- kept Poisson predictions but used quasi-Poisson to interpret frequency effects more honestly
- kept the raw Gamma baseline but added a capped-loss sensitivity to separate body-of-loss signal from extreme-tail noise
- rewrote the notebook around the actual modeling sequence so the reader can see the problem, the fix, and the practical implication in order

# What The Results Mean In Practice

I am not treating this as a final customer pricing model. What this notebook produces is expected claim cost per policy-year. In practice, that means:

- a policyholder costs more when the model expects them to claim more often
- a policyholder also costs more when, conditional on a claim, the model expects their claims to cost more
- the cleanest and most stable story in this notebook is on claim frequency
- severity adds signal, but that part of the story should be described more cautiously because large claims move it around

For this dataset, the practical headline feels pretty simple:

- younger drivers are materially more expensive risks in expected-loss terms
- some vehicle or usage proxies like fuel type and brand group matter as well
- the final premium differences are driven by the product of frequency and severity, not by one coefficient in isolation

# Limitations

- This is a side-project GLM baseline, not a production tariff.
- I am reading the model as association-based, not causal.
- The severity model still reacts to tail behavior even after the sensitivity check.
- Claim amounts may reflect convention-based French claims handling, which affects how I should interpret severity.
- If I were taking this further, the next steps would be richer treatment of overdispersion, more formal tail handling, and stronger governance review.
