---
title: "French MTPL GLM vs LightGBM Comparison"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

source("mtpl_gbm_analysis.R")
results <- run_mtpl_gbm_analysis()

checks <- results$checks
frequency_comparison <- results$frequency_comparison
severity_comparison <- results$severity_comparison
pure_premium_comparison <- results$pure_premium_comparison

get_check <- function(name) {
  unname(checks$value[checks$metric == name])
}

format_eur <- function(x) {
  paste0("EUR ", format(round(x, 0), big.mark = ",", scientific = FALSE))
}

format_pct <- function(x) {
  sprintf("%.2f%%", 100 * x)
}

round_table <- function(df, digits = 4) {
  df <- as.data.frame(df)
  numeric_cols <- vapply(df, is.numeric, logical(1))
  df[numeric_cols] <- lapply(df[numeric_cols], round, digits)
  df
}

freq_gbm_gini <- frequency_comparison$pricing_lift_gini[frequency_comparison$model == "LightGBM"]
freq_glm_gini <- frequency_comparison$pricing_lift_gini[frequency_comparison$model == "GLM"]

pure_gbm_gini <- pure_premium_comparison$pricing_lift_gini[
  pure_premium_comparison$model == "LightGBM" &
    pure_premium_comparison$target == "Raw severity pure premium"
]
pure_glm_gini <- pure_premium_comparison$pricing_lift_gini[
  pure_premium_comparison$model == "GLM" &
    pure_premium_comparison$target == "Raw severity pure premium"
]

comparison_row <- function(df, model, target = NULL) {
  if (is.null(target)) {
    df[df$model == model, , drop = FALSE][1, , drop = FALSE]
  } else {
    df[df$model == model & df$target == target, , drop = FALSE][1, , drop = FALSE]
  }
}

pct <- function(x) sprintf("%.1f%%", 100 * x)
num <- function(x, digits = 3) format(round(x, digits), big.mark = ",", scientific = FALSE)

freq_glm <- comparison_row(frequency_comparison, "GLM")
freq_gbm <- comparison_row(frequency_comparison, "LightGBM")
freq_gini_delta <- freq_gbm$pricing_lift_gini - freq_glm$pricing_lift_gini
freq_gini_rel_delta <- freq_gini_delta / freq_glm$pricing_lift_gini
freq_glm_claim_gap <- (freq_glm$predicted_claims - freq_glm$observed_claims) / freq_glm$observed_claims
freq_gbm_claim_gap <- (freq_gbm$predicted_claims - freq_gbm$observed_claims) / freq_gbm$observed_claims

raw_sev_glm <- comparison_row(severity_comparison, "GLM", "Raw severity")
raw_sev_gbm <- comparison_row(severity_comparison, "LightGBM", "Raw severity")
capped_sev_glm <- comparison_row(severity_comparison, "GLM", "Capped severity")
capped_sev_gbm <- comparison_row(severity_comparison, "LightGBM", "Capped severity")
raw_sev_glm_gap <- (raw_sev_glm$predicted_mean - raw_sev_glm$observed_mean) / raw_sev_glm$observed_mean
raw_sev_gbm_gap <- (raw_sev_gbm$predicted_mean - raw_sev_gbm$observed_mean) / raw_sev_gbm$observed_mean
capped_sev_glm_gap <- (capped_sev_glm$predicted_mean - capped_sev_glm$observed_mean) / capped_sev_glm$observed_mean
capped_sev_gbm_gap <- (capped_sev_gbm$predicted_mean - capped_sev_gbm$observed_mean) / capped_sev_gbm$observed_mean
raw_sev_mae_improvement <- (raw_sev_glm$mae - raw_sev_gbm$mae) / raw_sev_glm$mae
capped_sev_mae_improvement <- (capped_sev_glm$mae - capped_sev_gbm$mae) / capped_sev_glm$mae

raw_pp_glm <- comparison_row(pure_premium_comparison, "GLM", "Raw severity pure premium")
raw_pp_gbm <- comparison_row(pure_premium_comparison, "LightGBM", "Raw severity pure premium")
capped_pp_glm <- comparison_row(pure_premium_comparison, "GLM", "Capped severity pure premium")
capped_pp_gbm <- comparison_row(pure_premium_comparison, "LightGBM", "Capped severity pure premium")
raw_pp_gini_delta <- raw_pp_gbm$pricing_lift_gini - raw_pp_glm$pricing_lift_gini
raw_pp_gini_rel_delta <- raw_pp_gini_delta / raw_pp_glm$pricing_lift_gini
raw_pp_glm_gap <- (raw_pp_glm$predicted_loss_cost - raw_pp_glm$observed_loss_cost) / raw_pp_glm$observed_loss_cost
raw_pp_gbm_gap <- (raw_pp_gbm$predicted_loss_cost - raw_pp_gbm$observed_loss_cost) / raw_pp_gbm$observed_loss_cost
capped_pp_glm_gap <- (capped_pp_glm$predicted_loss_cost - capped_pp_glm$observed_loss_cost) / capped_pp_glm$observed_loss_cost
capped_pp_gbm_gap <- (capped_pp_gbm$predicted_loss_cost - capped_pp_gbm$observed_loss_cost) / capped_pp_gbm$observed_loss_cost

business_summary <- data.frame(
  question = c(
    "Frequency calibration",
    "Frequency ranking",
    "Severity calibration",
    "Severity error",
    "Pure premium ranking",
    "Pure premium calibration"
  ),
  result = c(
    paste0("GLM missed holdout claims by ", pct(freq_glm_claim_gap), "; LightGBM missed by ", pct(freq_gbm_claim_gap), "."),
    paste0("LightGBM frequency Gini improved by ", num(freq_gini_delta, 4), " points, or ", pct(freq_gini_rel_delta), "."),
    paste0("Raw severity remains weak: GLM mean gap ", pct(raw_sev_glm_gap), "; LightGBM mean gap ", pct(raw_sev_gbm_gap), "."),
    paste0("LightGBM lowered raw severity MAE by ", pct(raw_sev_mae_improvement), " and capped severity MAE by ", pct(capped_sev_mae_improvement), "."),
    paste0("LightGBM raw pure premium Gini improved by ", num(raw_pp_gini_delta, 4), " points, or ", pct(raw_pp_gini_rel_delta), "."),
    paste0("Raw pure premium is undercalibrated: GLM gap ", pct(raw_pp_glm_gap), "; LightGBM gap ", pct(raw_pp_gbm_gap), ".")
  ),
  implication = c(
    "The GLM remains the cleaner aggregate frequency benchmark.",
    "LightGBM finds additional segmentation signal in claim frequency.",
    "Neither model solves large-loss volatility; severity needs cautious interpretation.",
    "LightGBM helps error modestly, especially in the body of the severity distribution.",
    "The challenger is materially better at ordering policies by observed loss cost.",
    "LightGBM needs calibration before being used as a direct pricing level."
  ),
  row.names = NULL
)
```

# Executive Result

The actuarial objective is to estimate expected annual loss cost:

```text
expected annual pure premium = expected annual claim frequency * expected claim severity
```

The GLM is the benchmark. LightGBM is the challenger. The useful question is not whether the boosted model produces a nicer metric in isolation; it is whether it improves risk segmentation, calibration, or interaction discovery enough to justify the extra complexity.

On this holdout split, the useful result is mixed but informative:

- the GLM remains excellent on aggregate frequency calibration: it predicted `r num(freq_glm$predicted_claims, 1)` claims versus `r num(freq_glm$observed_claims, 0)` observed, a `r pct(freq_glm_claim_gap)` gap
- LightGBM is slightly worse on aggregate frequency calibration, predicting `r num(freq_gbm$predicted_claims, 1)` claims, but it improves frequency ranking from Gini `r round(freq_glm_gini, 4)` to `r round(freq_gbm_gini, 4)`
- raw pure premium Gini improves from `r round(pure_glm_gini, 4)` for the GLM to `r round(pure_gbm_gini, 4)` for LightGBM, a `r pct(raw_pp_gini_rel_delta)` relative lift
- LightGBM underpredicts aggregate pure premium more than the GLM, so its value is stronger as a segmentation and feature-discovery model than as an unadjusted pricing level
- the severity cap used for the sensitivity view is `r format_eur(results$severity_cap)`

```{r executive-model-value}
knitr::kable(results$model_value)
```

```{r business-summary}
knitr::kable(business_summary)
```

# Data And Split

This comparison reuses the same policy-level train/holdout split from the GLM project. That matters because the goal is a fair challenger comparison, not a new experiment with a different split.

```{r data-checks}
knitr::kable(round_table(checks))
```

The leakage checks should be zero. The observed-loss reconciliation gap should also be zero because pure premium validation depends on matching policy-level losses back to the severity file.

# Modeling Setup

The GLM benchmark stays actuarially classical:

- Poisson frequency with an exposure offset
- Gamma severity on positive claim rows
- pure premium derived as frequency times severity

The LightGBM challenger mirrors that structure:

- Poisson objective on annual claim frequency, weighted by exposure
- Gamma objective for raw severity
- Gamma objective for 99.5% capped severity
- pure premium derived from the frequency model and severity model

The holdout set is not used for tuning. LightGBM uses a small validation split inside the training set with early stopping.

```{r tuning-results}
tuning_all <- rbind(
  cbind(component = "Frequency", results$tuning$frequency),
  cbind(component = "Raw severity", results$tuning$severity),
  cbind(component = "Capped severity", results$tuning$capped_severity)
)
knitr::kable(round_table(tuning_all))
```

# Frequency

Frequency is usually the cleanest place for this dataset to show signal. The key actuarial checks are aggregate calibration, error on annual frequency, and whether the model ranks higher-frequency risks better.

```{r frequency-comparison}
knitr::kable(round_table(frequency_comparison))
```

The frequency result is the best argument for the GBM challenger. The GLM is almost perfectly calibrated in aggregate, missing holdout claim count by only `r pct(freq_glm_claim_gap)`. LightGBM misses by `r pct(freq_gbm_claim_gap)`, so it does not beat the GLM on total claim count. But LightGBM improves the ranking Gini by `r num(freq_gini_delta, 4)` points, or `r pct(freq_gini_rel_delta)` relative to the GLM. In actuarial terms, that means the GBM is better at sorting policies into higher- and lower-frequency risk groups, even though the GLM is already strong for aggregate expected claim count.

```{r frequency-decile-plot}
freq_deciles <- results$frequency_deciles
ylim <- range(c(freq_deciles$observed_annual_frequency, freq_deciles$predicted_annual_frequency), na.rm = TRUE)
plot(
  NULL,
  xlim = c(1, 10),
  ylim = ylim,
  xlab = "Predicted frequency decile",
  ylab = "Annual frequency",
  main = "Holdout Frequency Calibration"
)
for (model_name in unique(freq_deciles$model)) {
  rows <- freq_deciles[freq_deciles$model == model_name, ]
  line_col <- ifelse(model_name == "GLM", "gray40", "firebrick")
  points(rows$decile, rows$observed_annual_frequency, pch = 16, col = line_col)
  lines(rows$decile, rows$predicted_annual_frequency, type = "b", pch = 17, col = line_col)
}
legend("topleft", legend = unique(freq_deciles$model), col = c("gray40", "firebrick"), pch = 17, bty = "n")
```

# Severity

Severity is harder because claim amounts are heavy-tailed. The raw model is still useful to inspect, but the capped sensitivity is a more stable view of typical claim severity behavior.

```{r severity-comparison}
knitr::kable(round_table(severity_comparison))
```

The severity comparison is more cautious. On raw severity, the GLM predicts an average holdout severity of `r format_eur(raw_sev_glm$predicted_mean)` versus `r format_eur(raw_sev_glm$observed_mean)` observed, a `r pct(raw_sev_glm_gap)` gap. LightGBM predicts `r format_eur(raw_sev_gbm$predicted_mean)`, a `r pct(raw_sev_gbm_gap)` gap. That is worse aggregate calibration. However, LightGBM lowers raw severity MAE by `r pct(raw_sev_mae_improvement)`, and in the capped view its mean prediction is within `r pct(capped_sev_gbm_gap)` of observed capped severity.

My interpretation is that LightGBM finds some body-of-loss signal, but it does not solve the heavy-tail problem. For a portfolio project, that is still useful because it shows the right actuarial caution: the severity model should not be oversold.

```{r severity-decile-plot}
sev_deciles <- results$severity_deciles[results$severity_deciles$target == "Capped severity", ]
ylim <- range(c(sev_deciles$observed_mean, sev_deciles$predicted_mean), na.rm = TRUE)
plot(
  NULL,
  xlim = c(1, 10),
  ylim = ylim,
  xlab = "Predicted capped severity decile",
  ylab = "Mean capped severity",
  main = "Holdout Capped Severity Calibration"
)
for (model_name in unique(sev_deciles$model)) {
  rows <- sev_deciles[sev_deciles$model == model_name, ]
  line_col <- ifelse(model_name == "GLM", "gray40", "firebrick")
  points(rows$decile, rows$observed_mean, pch = 16, col = line_col)
  lines(rows$decile, rows$predicted_mean, type = "b", pch = 17, col = line_col)
}
legend("topleft", legend = unique(sev_deciles$model), col = c("gray40", "firebrick"), pch = 17, bty = "n")
```

# Pure Premium

Pure premium is the business target. At the policy level, observed annual loss cost is noisy because most policies have zero claim cost, but aggregate calibration and decile ranking are still useful.

```{r pure-premium-comparison}
knitr::kable(round_table(pure_premium_comparison))
```

Pure premium shows the tradeoff most clearly. The observed holdout loss cost is `r format_eur(raw_pp_glm$observed_loss_cost)` per exposure-year. The raw GLM pure premium predicts `r format_eur(raw_pp_glm$predicted_loss_cost)`, a `r pct(raw_pp_glm_gap)` gap. The raw LightGBM pure premium predicts `r format_eur(raw_pp_gbm$predicted_loss_cost)`, a larger `r pct(raw_pp_gbm_gap)` gap.

That looks bad if we only care about the unadjusted aggregate level. But the ranking story moves the other way: raw pure premium Gini improves from `r round(raw_pp_glm$pricing_lift_gini, 4)` to `r round(raw_pp_gbm$pricing_lift_gini, 4)`, a `r pct(raw_pp_gini_rel_delta)` relative improvement. This is exactly the distinction an insurer would care about. The GLM gives a more defensible base pricing level; the GBM gives stronger segmentation signal that could identify underpriced and overpriced groups.

```{r pure-premium-decile-plot}
pp_deciles <- results$pure_premium_deciles[results$pure_premium_deciles$target == "Raw severity pure premium", ]
ylim <- range(c(pp_deciles$observed_loss_cost, pp_deciles$predicted_loss_cost), na.rm = TRUE)
plot(
  NULL,
  xlim = c(1, 10),
  ylim = ylim,
  xlab = "Predicted pure premium decile",
  ylab = "Annual loss cost",
  main = "Holdout Pure Premium Calibration"
)
for (model_name in unique(pp_deciles$model)) {
  rows <- pp_deciles[pp_deciles$model == model_name, ]
  line_col <- ifelse(model_name == "GLM", "gray40", "firebrick")
  points(rows$decile, rows$observed_loss_cost, pch = 16, col = line_col)
  lines(rows$decile, rows$predicted_loss_cost, type = "b", pch = 17, col = line_col)
}
legend("topleft", legend = unique(pp_deciles$model), col = c("gray40", "firebrick"), pch = 17, bty = "n")
```

# Feature Importance

Feature importance is not a rate indication. I am using it as a diagnostic for which rating variables the challenger model used most heavily.

```{r feature-importance, results='asis'}
cat("## Frequency\n\n")
print(knitr::kable(head(round_table(results$importance$frequency), 10)))
cat("\n\n## Raw Severity\n\n")
print(knitr::kable(head(round_table(results$importance$severity), 10)))
cat("\n\n## Capped Severity\n\n")
print(knitr::kable(head(round_table(results$importance$capped_severity), 10)))
```

# Partial Dependence

Partial dependence gives a compact view of nonlinear patterns that a plain GLM may not capture directly.

```{r partial-dependence}
pdp <- results$pdp
for (target_name in unique(pdp$target)) {
  target_rows <- pdp[pdp$target == target_name, ]
  features <- unique(target_rows$feature)
  old_par <- par(mfrow = c(length(features), 1), mar = c(4, 4, 2, 1))
  for (feature_name in features) {
    rows <- target_rows[target_rows$feature == feature_name, ]
    plot(
      seq_len(nrow(rows)),
      rows$mean_prediction,
      type = "b",
      xaxt = "n",
      xlab = feature_name,
      ylab = "Mean prediction",
      main = paste(target_name, feature_name)
    )
    axis(1, at = seq_len(nrow(rows)), labels = rows$value, las = 2, cex.axis = 0.7)
  }
  par(old_par)
}
```

# Interaction Views

These tables are not a production interaction search. They are a sanity check for whether the boosted model is finding patterns worth considering as future GLM feature engineering.

```{r interactions, results='asis'}
cat("## DriverAge x CarAge\n\n")
print(knitr::kable(round_table(results$interactions$driver_car_age)))
cat("\n\n## Power x Brand\n\n")
print(knitr::kable(round_table(results$interactions$power_brand)))
cat("\n\n## Region x logDensity\n\n")
print(knitr::kable(round_table(results$interactions$region_density)))
```

# Conclusion

The final question is whether LightGBM adds actuarial value beyond the GLM. My answer is yes, but not as a clean replacement for the GLM.

The GLM is still the better benchmark model for a first actuarial pricing study. It is simpler, easier to explain, and very strong on aggregate frequency calibration. On holdout frequency, it predicts `r num(freq_glm$predicted_claims, 1)` claims against `r num(freq_glm$observed_claims, 0)` observed, while LightGBM predicts `r num(freq_gbm$predicted_claims, 1)`. That means I would not discard the GLM. It gives the clean baseline actuarial story: expected frequency, expected severity, and pure premium built from transparent model components.

The GBM adds value in a different way. Its main improvement is risk segmentation. Frequency Gini rises from `r round(freq_glm_gini, 4)` to `r round(freq_gbm_gini, 4)`, and raw pure premium Gini rises from `r round(pure_glm_gini, 4)` to `r round(pure_gbm_gini, 4)`. That is not a cosmetic improvement. It says the challenger is better at ordering risks by observed loss experience, which is useful for pricing review, underwriting triage, and discovering candidate interactions for a refined GLM.

The tradeoff is calibration. LightGBM underpredicts the aggregate raw pure premium more than the GLM: `r format_eur(raw_pp_gbm$predicted_loss_cost)` versus `r format_eur(raw_pp_gbm$observed_loss_cost)` observed, compared with `r format_eur(raw_pp_glm$predicted_loss_cost)` for the GLM. Severity remains the limiting piece. Raw severity is heavy-tailed, and although LightGBM reduces MAE, it does not produce a more credible raw aggregate severity level. The capped severity view is more stable, but that is a sensitivity analysis rather than a full tail model.

My recommendation is to keep the GLM as the primary model and use LightGBM as a challenger and diagnostic layer. For an insurer, the practical business impact would be:

- use the GLM to communicate the baseline technical premium structure
- use the GBM to find segments where the GLM may be smoothing over nonlinear or interaction effects
- investigate high-GBM/high-observed-loss segments as potential underpricing pockets
- investigate low-GBM/low-observed-loss segments as potential competitiveness opportunities
- recalibrate any GBM output before using it as an actual pricing level

For this side project, that is a good outcome. The GBM does uncover added signal, especially in ranking and segmentation, but the GLM remains sufficient as the core actuarial baseline. The strongest final story is not "GBM beats GLM." It is "GLM provides the credible benchmark, and GBM adds useful challenger insight that helps identify where the benchmark could be improved."
